---
title: "ATAC-seq exploration practical"
author: "Pierre-Luc"
date: "2023/04/21"
output: html_document
---

```{r}
suppressPackageStartupMessages({
  library(epiwraps)
  library(AnnotationHub)
  library(MotifDb)
  library(memes)
  library(universalmotif)
  library(ensembldb)
  library(ggplot2)
})

ah <- AnnotationHub(localHub=TRUE)
ensdb <- ah[["AH89211"]] # mouse ensembldb object
```


## Download the data

You can either start from the bam file:

```{r, eval=FALSE}
download.file("https://ethz-ins.org/content/w7/atac.chr19.bam", "atac.chr19.bam")
download.file("https://ethz-ins.org/content/w7/atac.chr19.bam.bai", "atac.chr19.bam.bai")
```

Alternatively, if you're on a slow computer, you can also download the tracks that we're generating below, and then skip that step:
```{r, eval=FALSE}
download.file("https://ethz-ins.org/content/w7/w7_bw.zip", "w7_bw.zip")
unzip("w7_bw.zip")
```


## Preparing tracks

```{r}
bam <- "atac.chr19.bam"

# create a track using all fragments, computing the coverage at each position (and force Ensembl naming)
epiwraps::bam2bw(bam, output_bw = "full_cov.bw", paired=TRUE, binWidth=10L, forceSeqlevelsStyle = "Ensembl")

# create a track using all fragments, but computing the number of cuts/insertion sites at each position
bam2bw(bam, output_bw = "full_cuts.bw", paired=TRUE, binWidth=1L, type="ends", 
                 shift=c(4L,-5L), forceSeqlevelsStyle = "Ensembl")

# create a track using only nucleosome-free fragments, computing coverage
bam2bw(bam, output_bw = "NF_cov.bw", paired=TRUE, binWidth=5L, minFragLength=30, 
       maxFragLength=120, forceSeqlevelsStyle = "Ensembl")

# create a track using only nucleosome-free fragments, the number of cuts/insertion sites at each position
bam2bw(bam, output_bw = "NF_cuts.bw", paired=TRUE, binWidth=1L, type="ends", extend=2L, minFragLength=30, 
       maxFragLength=120, shift=c(4L,-5L), forceSeqlevelsStyle = "Ensembl")

# create a track using only mono-nucleosome fragments, computing coverage
bam2bw(bam, output_bw = "mono_cov.bw", paired=TRUE, binWidth=10L, minFragLength=140,
       maxFragLength=220, forceSeqlevelsStyle = "Ensembl")

# create a track using only the (10bp) centers of mono-nucleosome fragments
bam2bw(bam, output_bw = "mono_centers.bw", paired=TRUE, binWidth=5L, minFragLength=140, shift=c(4L,-5L), 
       maxFragLength=220, type="center", extend=10L, forceSeqlevelsStyle = "Ensembl")


list.files()
```


## Obtaining the sites with a CTCF motif

```{r}
# get CTCF motif
motif <- MotifDb::query(MotifDb, c("CTCF","Mus"))[[1]]
motif2 <- convert_motifs(motif, class="TFBSTools-PFMatrix")
genome <- ah[["AH68356"]]
# get the sequence for chr19:
chr19 <- import(genome)["19"]

# find motif matches across chr19
moi <- motifmatchr::matchMotifs(motif2, chr19, out="positions", p.cutoff=1e-5)[[1]]
# convert to GRanges
moi <- as(setNames(moi,names(chr19)), "GRanges")
```

## Plotting signal around the motif occurences

```{r, fig.width=8, fig.height=4}
# we prepare the list of tracks
tracks <- c("Full coverage"="full_cov.bw", "NF coverage"="NF_cov.bw", "All cuts"="full_cuts.bw",
            "NF cuts"="NF_cuts.bw", "Mono cov"="mono_cov.bw", "Mono centers"="mono_centers.bw")

# note that we could also do something like this programmatically:
# tracks <- list.files(pattern="bw$")
# names(tracks) <- gsub("\\.bw","",basename(tracks))

# extract signals around the motif occurences
# we zoom in to 300bp around the motif centers, in windows of 5bp
sm <- signal2Matrix(tracks, moi, w=5, extend=300)

# plot the signals:
plotEnrichedHeatmaps(sm, trim=0.95)

# the right to do this, however, is using a background normalization
# this we can do for instance using:
nf <- getNormFactors(tracks, useSeqLevels="19", nwind=5000L)
# then we apply the normalization factors:
sm <- renormalizeSignalMatrices(sm, scaleFactors = nf)
# we subset to the top 1000 regions:
# sm.top <- head(sm[order(rowMeans(as.matrix(assays(sm)$enriched_score))),], 500)
plotEnrichedHeatmaps(sm, trim=0.95, minRowVal = 12, colors = c("white","darkred"))
```


## Plotting footprint

We could use our standard functions:

```{r}
d <- meltSignals(sm[,c("NF cuts", "Mono centers")])
ggplot(d, aes(position, mean, colour=sample)) + geom_line(size=1.2)
```

Or we could use a specialized function from the ATACseqQC package:

```{r}
library(ATACseqQC)
epiwraps::motifFootprint("atac.chr19.bam", motif, moi)
```

## Signal around TSS

We could also look at the pattern around the TSS

```{r}
# fetch the protein-coding TSS from chromosome 19
txs <- transcripts(ensdb, filter=TxBiotypeFilter("protein_coding"))
txs <- txs[seqnames(txs)=="19"]
tss <- promoters(txs, upstream=0, downstream=1)
tss.sm <- epiwraps::signal2Matrix(tracks, granges(tss), w=10, extend=1000)
tss.sm <- epiwraps::renormalizeSignalMatrices(tss.sm, scaleFactors = nf)
epiwraps::plotEnrichedHeatmaps(tss.sm, trim=0.95, colors=c("white","darkred"))
```



